Investigating soundscapes perception through acoustic scenes simulation

### 2018-06-26 v2/responseToReviewers.tex

\documentclass[10pt]{article}

\title{Reply to reviewers concerning submission BR-Org-18-034 "Investigating soundscapes perception through acoustic scenes simulation"}

@@

\begin{document}

\maketitle

As a preamble, we would like to thank the editor and the reviewer for their comments and suggestions. Following these comments, we made several changes to the article, which are summarized here. The next sections list our answers to each of the reviewer’s comments, with references to the revised manuscript (page, column, and paragraph) where appropriate.

\section{Answer to Academic Editor}

\item \emph{please follow APA (American Psychological Association) style guidelines (w.r.t. in-text citations, etc.).}

$\rightarrow$ The manuscript now follows the APA citation style. The text have been modified accordingly.

\item \emph{ Please avoid the over-use of acronyms as it makes reading of your article unnecessarily difficult (examples: P@5, ni, i, etc.). This sort of lab jargon is not transparent to the wider readership of BRM. }

$\rightarrow$ Those abbreviations are now in full text, except for some figures for the sake of clarity, in which case the acronyms are detailed in the caption.

\item \emph{I would also like to see comparisons between your approach and other state-of-the-art soundscape analyses (e.g. TAPESTREA, see comment from the reviewer below).}

$\rightarrow$ TODO

\item \emph{You might consider providing a link to a github site with examples of your stimuli. It is hard to imagine what these sound scenes sound like without some real examples.}

The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}. The link is now in the introduction. The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.

### 2018-06-26 lafaySimulatedScenePerception.tex
adressing editor comments

\documentclass[12pt]{elsarticle}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage[position=top]{subfig}
\usepackage{caption}
\usepackage{sectsty}
\usepackage{amsmath}
\usepackage{color}

@@

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments;
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments using \emph{Simscene} (see Section~\ref{sec:simulator}). Each of them has to create two sound environments: one ideal~/ pleasant, and the other non ideal~/ unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes. For that, a second group of subjects is asked to evaluate the pleasantness of each scene produced during (1.a), on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene;
\item to detect the presence of \emph{outliers} or ambiguous scenes. Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as reference. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the least pleasant i-scene remains more highly rated on average than the most pleasant ni-scene.
\end{enumerate}
\end{itemize}

@@

Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                              & Acronyms                   \\
\hline
Sound level                        & $L$                        \\
Sound level (events)               & $L(E)$                     \\
Sound level (textures)             & $L(T)$                     \\
Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
                                   &                            \\
ideal/pleasant                     & i                          \\
non-ideal/unpleasant               & ni                         \\
scene ideal/pleasant               & i-scene                    \\
scene non-ideal/unpleasant         & ni-scene                   \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the sensory experiments.}
\label{tab:acronyme}
\end{table}

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores. $\mathcal{A}_{subject}$ is computed for i-scenes and ni-scenes.

@@

\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes during the experiment 1.a. To do so, we consider if there exist significant differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.

@@

\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the i- and ni-scenes. The significance is measured with a two-sample Student test.

@@

where $\bigoplus$ is the \emph{exclusive-or} operator. Two scenes having similar source compositions will be close in such space. Using the Hamming distance allows us to take into account equally the presence and absence of classes. In order to measure the intrinsic ability of the space to discriminate between i- and ni-scenes, we use a ranking metric named the precision at rank $k$ ($P@k$). The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the proportion of $s_j$ scenes in the $k$ nearest neighbors of $s_i$ that share the same label as $s_i$. The $P@k$ is then the average of this ratio for all the items considered as search seeds.

@@

First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene. 4 ni-scenes do not fulfill that constraint: they are thus removed, together with their corresponding i-scenes. As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between i- and ni-scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.

@@

\begin{figure}[t]
		\myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_1}\label{fig:soundlevela}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_3}\label{fig:soundlevelb}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_5}\label{fig:soundlevelc}}\par
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_2}\label{fig:soundleveld}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_4}\label{fig:soundlevele}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
\end{figure}

@@

First, our analysis focuses on the sound levels. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$. There is a significant difference in terms of sound levels between i- and ni-scenes ($L$: $p<0.01$). This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).

@@

As expected, the sound level of the sources is indeed a pleasantness indicator, as the ni-scenes tend to be louder. This results is also an outcome of a large number of related works. We also notice that this difference of sound levels is significant for both events and textures.

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie~the events.

@@

However, level alone is not sufficient to fully differentiate between i- and ni-scenes. In fact, $20\%$ of the i-scenes have a sound levels higher than the lowest level of the ni-scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & all scenes                & i-scenes          & ni-scenes    \\
\hline
$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$)\\
$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$)\\
$L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$) \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and sound levels.}
\label{tab:corrStructA}
\end{table}

@@

In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated. Contrary to the previous test, we do not limit ourselves to a binary i-scenes \emph{vs.} ni-scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature. The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown in Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for i- and ni-scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for i-scenes.

@@

When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i-scenes tend to be softer than the ni-scenes, thus allowing us to extend artificially to the i-scenes the negative correlation observed for the ni-scenes.

@@

\begin{itemize}
\item allows us to differentiate between i- and ni-scenes,
\item characterizes precisely the perceived pleasantness for ni-scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
\item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (i-scene).
\end{itemize}

@@

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (i-scenes tend to be softer than ni-scenes). Thus, while sound event levels maintain a relative ability to predict the pleasantness of the ni-scenes, texture levels do not bring much information, whatever the type of environment is.

@@

Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (i or ni), and 2) each involves distinct, independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the i-scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness. We thus put forward the hypothesis that only the level of some of them can influence this perception. In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie we take an interest in the nature of the sources they are made of.

@@

We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (i or ni). For the 36 i and ni-scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.

@@

We observe a noticable difference in terms of class choices between the i- and ni-scenes. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the i-scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the ni-scenes. These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.

@@

To a certain extent, our results contradict this result. In fact,  Figure~\ref{fig:soundsourcea} shows that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes. Those results do not question the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes. However, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{truck} classes. On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.

@@

\begin{table}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked as decreasing order of V-test value, shown between parenthesis. $p\leq0.01$ for all sound markers.}
\label{tab:markers}
\end{table}

@@

We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between i- or ni-scenes. We now investigate whether some of the sound classes are specific to a given environment. For that purpose, the V-test detailed in  Section~\ref{sec:xp1_dataAna} is considered separately for each abstraction level. Results are presented in Table~\ref{tab:markers}.

@@

Regarding the sound events, 9 markers are identified for all abstraction levels. As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class. This latter result may be due to the socio-cultural background of the subjects who are mostly European citizens. In fact, according to Schafer, a sound that is identified by a person as being an important element of his/her environment, is well accepted. Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).

@@

Regarding the textures, 5 markers are identified. For the i-scenes, those classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}). The marker classes for the ni-scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).

>>>>

\documentclass[12pt]{elsarticle}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage[position=top]{subfig}
\usepackage{caption}
\usepackage{sectsty}
\usepackage{amsmath}
\usepackage{color}
%\usepackage[natbibapa]{apacite}

@@

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments using \emph{Simscene} (see Section~\ref{sec:simulator}). Each of them has to create two sound environments: one ideal~/ pleasant, and the other non ideal~/ unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes. For that, a second group of subjects is asked to evaluate the pleasantness of each scene produced during (1.a), on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (ideal or non ideal) thanks to a finer quantification of the pleasantness of the scene;
\item to detect the presence of \emph{outliers} or ambiguous scenes. Indeed, throughout our analyses, the predefined hedonic properties of the scenes (ideal or not) are used as reference. We thus need to ensure beforehand that no ambiguity exists between extreme cases of ideal and non ideal scenes, \ie~that the least pleasant ideal scene remains more highly rated on average than the most pleasant non ideal scene.
\end{enumerate}
\end{itemize}

@@

Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.  The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                              & Acronyms                   \\
\hline
Sound level                        & $L$                        \\
Sound level (events)               & $L(E)$                     \\
Sound level (textures)             & $L(T)$                     \\
Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
%                                    &                            \\
% ideal/pleasant                     & i                          \\
% non-ideal/unpleasant               & ni                         \\
% scene ideal/pleasant               & i-scene                    \\
% scene non-ideal/unpleasant         & ni-scene                   \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the sensory experiments.}
\label{tab:acronyme}
\end{table}

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores. $\mathcal{A}_{subject}$ is computed for ideal scenes and non ideal scenes.

@@

\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and non ideal scenes during the experiment 1.a. To do so, we consider if there exist significant differences for the i- and non ideal scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.

@@

\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the i- and non ideal scenes. The significance is measured with a two-sample Student test.

@@

where $\bigoplus$ is the \emph{exclusive-or} operator. Two scenes having similar source compositions will be close in such space. Using the Hamming distance allows us to take into account equally the presence and absence of classes. In order to measure the intrinsic ability of the space to discriminate between i- and non ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$). The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the proportion of $s_j$ scenes in the $k$ nearest neighbors of $s_i$ that share the same label as $s_i$. The $P@k$ is then the average of this ratio for all the items considered as search seeds.

@@

First, in order to ensure the coherence of the data, we check that none of the non ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene. 4 non ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes. As a consequence, 36 ideal scenes and 36 non ideal scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between i- and non ideal scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non ideal scenes.

@@

\begin{figure}[t]
		\myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_1}\label{fig:soundlevela}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_3}\label{fig:soundlevelb}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_5}\label{fig:soundlevelc}}\par
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_2}\label{fig:soundleveld}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_4}\label{fig:soundlevele}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (i: ideal, ni: non ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
\end{figure}

@@

First, our analysis focuses on the sound levels. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$. There is a significant difference in terms of sound levels between ideal and non ideal scenes ($L$: $p<0.01$). This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).

@@

As expected, the sound level of the sources is indeed a pleasantness indicator, as the non ideal scenes tend to be louder. This results is also an outcome of a large number of related works. We also notice that this difference of sound levels is significant for both events and textures.

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non ideal scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie~the events.

@@

However, level alone is not sufficient to fully differentiate between ideal and non ideal scenes. In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & all scenes                & ideal scenes          & non ideal scenes    \\
\hline
$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$)\\
$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$)\\
$L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$) \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and sound levels.}
\label{tab:corrStructA}
\end{table}

@@

In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated. Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non ideal scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature. The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown in Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non ideal scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for non ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for ideal scenes.

@@

When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the ideal scenes tend to be softer than the non ideal scenes, thus allowing us to extend artificially to the ideal scenes the negative correlation observed for the non ideal scenes.

@@

\begin{itemize}
\item allows us to differentiate between i- and non ideal scenes,
\item characterizes precisely the perceived pleasantness for non ideal scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
\item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (ideal scene).
\end{itemize}

@@

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non ideal scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non ideal scenes). Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non ideal scenes, texture levels do not bring much information, whatever the type of environment is.

@@

Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (i or ni), and 2) each involves distinct, independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the ideal scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness. We thus put forward the hypothesis that only the level of some of them can influence this perception. In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie we take an interest in the nature of the sources they are made of.

@@

We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (i or ni). For the 36 i and non ideal scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.

@@

We observe a noticeable difference in terms of class choices between the i- and non ideal scenes. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes. These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.

@@

To a certain extent, our results contradict this result. In fact,  Figure~\ref{fig:soundsourcea} shows that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the ideal scenes, and for $42\%$ of the non ideal scenes. Those results do not question the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the ideal scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes. However, \emph{public transport} classes are also strongly present in the non ideal scenes, for instance more than \emph{light vehicle} or \emph{truck} classes. On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.

@@

\begin{table}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & ideal scenes & non ideal scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & ideal scenes & non ideal scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked as decreasing order of V-test value, shown between parenthesis. $p\leq0.01$ for all sound markers.}
\label{tab:markers}
\end{table}

@@

We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between i- or non ideal scenes. We now investigate whether some of the sound classes are specific to a given environment. For that purpose, the V-test detailed in  Section~\ref{sec:xp1_dataAna} is considered separately for each abstraction level. Results are presented in Table~\ref{tab:markers}.

@@

Regarding the sound events, 9 markers are identified for all abstraction levels. As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are ideal scenes markers as well as the \emph{church bell} class. This latter result may be due to the socio-cultural background of the subjects who are mostly European citizens. In fact, according to Schafer, a sound that is identified by a person as being an important element of his/her environment, is well accepted. Sound markers of non ideal scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).

@@

Regarding the textures, 5 markers are identified. For the ideal scenes, those classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}). The marker classes for the non ideal scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).
